Subgroup detection in linear growth curve models with GLMM trees

### 2023-08-15 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
small touch-ups, esp grey -> gray
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3269 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c72b022b6fab2678f0fbbc00f4247c7da9e7535a

For example, Figure~\ref{fig:global_trajectories} shows trajectories of science knowledge and skills from 250 children who were assessed at three timepoints throughout grades 3 through 8\footnote{Further details on the source of these data and scaling of the response variable are provided in Study~III.}. The grey lines depict individual trajectories, while the red line depicts the estimated average trajectory. Although the red line describes the average trajectory well, the grey lines indicate substantial intra-individual variability in average levels and growth over time. A common research aim might be to identify which covariates can explain or predict this heterogeneity.

\begin{figure}%
\caption{Growth curves of science ability.}
\begin{subfigure}{0.65\textwidth}%
<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
lines(x = c(min(data$months), max(data$months)), 
     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
     lwd = 3, col = 2)
@
\\\footnotesize{\textit{Note.} Grey lines depict observed individual trajectories. The red line depicts the average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.}
\end{subfigure}%
\label{fig:global_trajectories}
\end{figure}%

\begin{figure}%
\caption{Partitioned growth curves of science abilities.}
\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=9, width=12, echo=FALSE>>=
LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
testdat$CHILDID <- factor(testdat$CHILDID)
testdat$GENDER <- factor(testdat$GENDER)
testdat$asmtmm <- factor(testdat$asmtmm)
## Remove obs with levels of race in test that are not in train
levs <- unique(testdat$RACE) %in% unique(data$RACE)
if (!all(levs)) {
  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
}
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
                                        xaxis.labs = c(0, 12, 36, 60)),
     gp = gpar(cex = 1.5))
icc_c <- as.data.frame(VarCorr(lmm_tree_c))
lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
lmm_preds <- predict(LMM, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Grey lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercepts with respect to individuals.}
\end{subfigure}%
\label{fig:LMM_tree_c}
\end{figure}%

@@

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_c} shows an example tree, which identified socio-economic status (SES) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of these children, assessed at baseline. Four subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope (see Table~\ref{tab:local_coefs}). For children in the group with lower levels of SES, the intercept is also lowest, indicating lower science ability. For intermediate levels of SES (between -0.41 and 1.23), there is also an effect of internalizing problems, where children with higher internalizing problems have a lower intercept. Highest science ability intercepts are observed in the group with highest SES. It appears that nodes with higher intercepts also tend to have higher slopes, indicating that children with relatively higher ability also show larger ability gains over time.

@@

\begin{figure}%
\caption{GLMM tree estimated by initializing with zero random effects.}
<<fig=TRUE, height=7, width=12, echo=FALSE, out.width=12>>=
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
icc <- as.data.frame(VarCorr(lmm_tree))
lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability. Grey lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree}
\end{figure}%

\begin{figure}%
\caption{GLMM tree initialized by estimating the random effects.}
<<fig=TRUE, height=9, width=15, echo=FALSE>>=
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
icc_r <- as.data.frame(VarCorr(lmm_tree_r))
lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Grey lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree_r}
\end{figure}%

@@

breaks <- c(0, 1, 3, 7, 15, 20, 25.5, 32)
labels <- c("0", "1", "3", "7", "15", "MAD", "SD", "M")
lim <- c(0, 30)
plot1 <- ggplot(tmp2) +
  geom_boxplot(aes(x=random, y=tree_size), alpha = .5, width = .6, coef = NULL) +
  geom_count(aes(x=random, y=tree_size), col = "gray", alpha = .75) +
  geom_hline(yintercept=3, col = "darkgray") +
  geom_text(data = summ_stats, aes(x=random, y=y, label = all),
            size = 3, col = "gray44") +
  scale_y_continuous(trans = pseudo_log_trans(), 
                     breaks = breaks, lim = lim, labels = labels) +
  facet_grid(~ method, scales = "free", space = "free") +
  labs(x = "", y = "no. of splits") + theme(legend.position = "none") +
  scale_x_discrete(labels = scales::parse_format()) +
  coord_cartesian(clip = 'off')
plot1
@
\end{subfigure}%
\vspace{-0.5cm}%
{
\footnotesize \textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= Mean absolute deviation from true tree size. Grey circles represent counts, dark gray horizontal lines indicate true number of splits~3. Distances on the $y$-axis are on the log scale; $x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.}
\label{fig:tree_sizes}
\end{figure}

@@

## Add benchmark LMMs
load("LMM MSEs, N = 250, Math.RDa")
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 1] <- format(colMeans(MSEs), digits = 4)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 2] <- round(sapply(MSEs, sd), digits = 3)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 3] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_math)

load("LMM MSEs, N = 250, Reading.RDa")
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 4] <- format(colMeans(MSEs), digits = 4)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 5] <- round(sapply(MSEs, sd), digits = 3)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 6] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_read)

load("LMM MSEs, N = 250, Science.RDa")
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 7] <- format(colMeans(MSEs), digits = 4)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 8] <- round(sapply(MSEs, sd), digits = 3)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 9] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_scie)

>>>>

For example, Figure~\ref{fig:global_trajectories} shows trajectories of science knowledge and skills from 250 children who were assessed at three timepoints throughout grades 3 through 8\footnote{Further details on the source of these data and scaling of the response variable are provided in Study~III.}. The gray lines depict individual trajectories, while the red line depicts the estimated average trajectory. Although the red line describes the average trajectory well, the gray lines indicate substantial intra-individual variability in average levels and growth over time. A common research aim might be to identify which covariates can explain or predict this heterogeneity.

\begin{figure}%
\caption{Growth curves of science ability.}
\begin{subfigure}{0.65\textwidth}%
<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
lines(x = c(min(data$months), max(data$months)), 
     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
     lwd = 3, col = 2)
@
\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories. The red line depicts the average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.}
\end{subfigure}%
\label{fig:global_trajectories}
\end{figure}%

\begin{figure}%
\caption{Partitioned growth curves of science abilities.}
\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=9, width=12, echo=FALSE>>=
LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
testdat$CHILDID <- factor(testdat$CHILDID)
testdat$GENDER <- factor(testdat$GENDER)
testdat$asmtmm <- factor(testdat$asmtmm)
## Remove obs with levels of race in test that are not in train
levs <- unique(testdat$RACE) %in% unique(data$RACE)
if (!all(levs)) {
  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
}
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
                                        xaxis.labs = c(0, 12, 36, 60)),
     gp = gpar(cex = 1.5))
icc_c <- as.data.frame(VarCorr(lmm_tree_c))
lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
lmm_preds <- predict(LMM, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.}
\end{subfigure}%
\label{fig:LMM_tree_c}
\end{figure}%

@@

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_c} shows an example tree, which identified socio-economic status (SES) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of these children, assessed at baseline. Four subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope (see Table~\ref{tab:local_coefs}). For children in the group with lower levels of SES, the intercept is also lowest, indicating lower science ability. For intermediate levels of SES (between $-0.41$ and $1.23$), there is also an effect of internalizing problems, where children with higher internalizing problems have a lower intercept. Highest science ability intercepts are observed in the group with highest SES. It appears that nodes with higher intercepts also tend to have higher slopes, indicating that children with relatively higher ability also show larger ability gains over time.

@@

\begin{figure}%
\caption{GLMM tree estimated by initializing with zero random effects.}
<<fig=TRUE, height=7, width=12, echo=FALSE, out.width=12>>=
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
icc <- as.data.frame(VarCorr(lmm_tree))
lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree}
\end{figure}%

\begin{figure}%
\caption{GLMM tree initialized by estimating the random effects.}
<<fig=TRUE, height=9, width=15, echo=FALSE>>=
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
icc_r <- as.data.frame(VarCorr(lmm_tree_r))
lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree_r}
\end{figure}%

@@

breaks <- c(0, 1, 3, 7, 15, 20, 25.5, 32)
labels <- c("0", "1", "3", "7", "15", "MAD", "SD", "M")
lim <- c(0, 30)
plot1 <- ggplot(tmp2) +
  geom_boxplot(aes(x=random, y=tree_size), alpha = .5, width = .6, coef = NULL) +
  geom_count(aes(x=random, y=tree_size), col = "gray", alpha = .75) +
  geom_hline(yintercept=3, col = "darkgray") +
  geom_text(data = summ_stats, aes(x=random, y=y, label = all),
            size = 3, col = "gray44") +
  scale_y_continuous(trans = pseudo_log_trans(), 
                     breaks = breaks, lim = lim, labels = labels) +
  facet_grid(~ method, scales = "free", space = "free") +
  labs(x = "", y = "no. of splits") + theme(legend.position = "none") +
  scale_x_discrete(labels = scales::parse_format()) +
  coord_cartesian(clip = 'off')
plot1
@
\end{subfigure}%
\vspace{-0.5cm}%
{
\footnotesize \textit{Note.} M~= mean number of splits; SD~= standard deviation of the number of splits; MAD~= Mean absolute deviation from true tree size. Gray circles represent counts, dark gray horizontal lines indicate true number of splits~3. Distances on the $y$-axis are on the log scale; $x$-axis labels indicate whether variances of the random intercept and slope were fixed to 0 or freely estimated.}
\label{fig:tree_sizes}
\end{figure}

@@

## Add benchmark LMMs
load("LMM MSEs, N = 250, Math.Rda")
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 1] <- format(colMeans(MSEs), digits = 4)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 2] <- round(sapply(MSEs, sd), digits = 3)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 3] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_math)

load("LMM MSEs, N = 250, Reading.Rda")
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 4] <- format(colMeans(MSEs), digits = 4)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 5] <- round(sapply(MSEs, sd), digits = 3)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 6] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_read)

load("LMM MSEs, N = 250, Science.Rda")
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 7] <- format(colMeans(MSEs), digits = 4)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 8] <- round(sapply(MSEs, sd), digits = 3)
MSE_df[c("LMM$^i$", "LMM$^{i,s}$"), 9] <- sprintf("%.3f", 1 - colMeans(MSEs)/var_scie)
